# pre-class
```{r}
#setup
install.packages("gapminder")
library(gapminder)
library(dplyr)
gapminder
```
Make sure you commit this often with meaningfull messages. 

gapminder
### Getting Started


We will work with the dataset called [gapminder](https://github.com/jennybc/gapminder), this is a cleaned up version from [Gapminder Data](http://www.gapminder.org/data/). Gapminder contains a lot of great data on all of the nations of the world. We first need to install the gapminder package in R. 

### Pre-Class Problems

Use **dplyr** functions to address the following questions:

1. How many unique countries are represented per continent?
There are 142 countries overall. 2 countries in Oceania are represented, 30 in Europe, 33 in Asia, 25 in the Americas, and 52 in Africa.
```{r}
    uni_country_tbl<-gapminder%>%
    distinct(country, continent)%>%
    arrange(continent)

#filter the amount of observations by each continent 
uni_country_tbl%>%
  filter(continent=="Africa")
uni_country_tbl%>%
  filter(continent=="Americas")
uni_country_tbl%>%
  filter(continent=="Asia")
uni_country_tbl%>%
  filter(continent=="Europe")
uni_country_tbl%>%
  filter(continent=="Oceania")
```
  
2. Which European nation had the lowest GDP per capita in 1997? 
Albania had the lowest GDP per capita in 1997 at 3193.
```{r}
gapminder%>%
    filter(continent=="Europe", year=="1997")%>%
    arrange(gdpPercap)
```
3. According to the data available, what was the average life expectancy across each continent in the 1980s?
Average life expectancy for Africa was 52.5 years, for Asia was 63.7 years, for the Americas was 67.2 years, for Europe was 73.2 years, and for Oceania was 74.8 years.
```{r}
gapminder %>%
    filter(year>=1980 & year<=1989)%>%
    group_by(continent)%>%
    summarise(avglifeexp = mean(lifeExp))%>%
    arrange(avglifeexp)
```
4. What 5 countries have the highest total GDP over all years combined?
The US, Japan, China, Germany, and the UK all had the highest total GDP across all years combined.
```{r}
gapminder%>%
    group_by(country)%>%
    mutate(gdp=gdpPercap*pop)%>%
    summarise(totalGDP=sum(gdp))%>%
    arrange(desc(totalGDP))
```
5. What countries and years had life expectancies of _at least_ 80 years? _N.b. only output the columns of interest: country, life expectancy and year (in that order)._
These are the following countries and years:
country          lifeExp  year
   <fct>              <dbl> <int>
 1 Australia           80.4  2002
 2 Australia           81.2  2007
 3 Canada              80.7  2007
 4 France              80.7  2007
 5 Hong Kong, China    80    1997
 6 Hong Kong, China    81.5  2002
 7 Hong Kong, China    82.2  2007
 8 Iceland             80.5  2002
 9 Iceland             81.8  2007
10 Israel              80.7  2007
11 Italy               80.2  2002
12 Italy               80.5  2007
13 Japan               80.7  1997
14 Japan               82    2002
15 Japan               82.6  2007
16 New Zealand         80.2  2007
17 Norway              80.2  2007
18 Spain               80.9  2007
19 Sweden              80.0  2002
20 Sweden              80.9  2007
21 Switzerland         80.6  2002
22 Switzerland         81.7  2007
```{r}
gapminder%>%
  filter(lifeExp>=80)%>%
  select(country, lifeExp, year)%>%
  print(n=22)
```

6. What 10 countries have the strongest correlation (in either direction) between life expectancy and per capita GDP?
France, Austria, Belgium, Norway, and Oman have the strongest correlation between life expectancy and per capita GDP.

```{r}
gapminder%>%
    group_by(country)%>%
    #since "strongest", need absolute value
    summarize(corr = abs(cor(lifeExp, gdpPercap)))%>%
    arrange(desc(corr))
```
7. Which combinations of continent (besides Asia) and year have the highest average population across all countries? _N.b. your output should include all results sorted by highest average population_. With what you already know, this one may stump you. See [this Q&A](http://stackoverflow.com/q/27207963/654296) for how to `ungroup` before `arrange`ing. This also [behaves differently in more recent versions of dplyr](https://github.com/hadley/dplyr/releases/tag/v0.5.0).
The five highest combinations are all in the Americas in the years 2007, 2002, 1997, 1992, and 1987.
 continent  year    avgpop
   <fct>     <int>     <dbl>
 1 Americas   2007 35954847.
 2 Americas   2002 33990910.
 3 Americas   1997 31876016.
 4 Americas   1992 29570964.
 5 Americas   1987 27310159.
```{r}
gapminder%>%
  group_by(continent, year)%>%
  filter(continent != "Asia")%>%
  summarise(avgpop=mean(pop))%>%
  arrange(desc(avgpop))
```
8. Which three countries have had the most consistent population estimates (i.e. lowest standard deviation) across the years of available data? 
These countries were Sao Tome and Principe, Iceland, and Montenegro.
   country                 sdpop
   <fct>                   <dbl>
 1 Sao Tome and Principe  45906.
 2 Iceland                48542.
 3 Montenegro             99738
```{r}
gapminder%>%
  group_by(country)%>%
  summarise(sdpop = sd(pop))%>%
  arrange(sdpop)
```
9. Subset **gm** to only include observations from 1992 and store the results as **gm1992**. What kind of object is this?
GM1992 is a list, with a tbl, tbl_df, and data.frame.
```{r}
gm1992<- gapminder%>%
  filter(year==1992)
  typeof(gm1992)
  class(gm1992)
```
10. Which observations indicate that the population of a country has *decreased* from the previous year **and** the life expectancy has *increased* from the previous year? See [the vignette on window functions](https://cran.r-project.org/web/packages/dplyr/vignettes/window-functions.html).

Here are the 36 observations for which this is true.

 country                continent  year lifeExp      pop gdpPercap
   <fct>                  <fct>     <int>   <dbl>    <int>     <dbl>
 1 Afghanistan            Asia       1982    39.9 12881816      978.
 2 Bosnia and Herzegovina Europe     1992    72.2  4256013     2547.
 3 Bosnia and Herzegovina Europe     1997    73.2  3607000     4766.
 4 Bulgaria               Europe     2002    72.1  7661799     7697.
 5 Bulgaria               Europe     2007    73.0  7322858    10681.
 6 Croatia                Europe     1997    73.7  4444595     9876.
 7 Czech Republic         Europe     1997    74.0 10300707    16049.
 8 Czech Republic         Europe     2002    75.5 10256295    17596.
 9 Czech Republic         Europe     2007    76.5 10228744    22833.
10 Equatorial Guinea      Africa     1977    42.0   192675      959.
11 Germany                Europe     1977    72.5 78160773    20513.
12 Germany                Europe     1987    74.8 77718298    24639.
13 Guinea-Bissau          Africa     1967    35.5   601287      716.
14 Hungary                Europe     1987    69.6 10612740    12986.
15 Hungary                Europe     1997    71.0 10244684    11713.
16 Hungary                Europe     2002    72.6 10083313    14844.
17 Hungary                Europe     2007    73.3  9956108    18009.
18 Ireland                Europe     1957    68.9  2878220     5599.
19 Ireland                Europe     1962    70.3  2830000     6632.
20 Kuwait                 Asia       1992    75.2  1418095    34933.
21 Lebanon                Asia       1982    67.0  3086876     7641.
22 Montenegro             Europe     2007    74.5   684736     9254.
23 Poland                 Europe     2002    74.7 38625976    12002.
24 Poland                 Europe     2007    75.6 38518241    15390.
25 Portugal               Europe     1972    69.3  8970450     9022.
26 Romania                Europe     1997    69.7 22562458     7347.
27 Romania                Europe     2002    71.3 22404337     7885.
28 Romania                Europe     2007    72.5 22276056    10808.
29 Rwanda                 Africa     1997    36.1  7212583      590.
30 Serbia                 Europe     2002    73.2 10111559     7236.
31 Slovenia               Europe     2002    76.7  2011497    20660.
32 Slovenia               Europe     2007    77.9  2009245    25768.
33 Switzerland            Europe     1977    75.4  6316424    26982.
34 Trinidad and Tobago    Americas   1992    69.9  1183669     7371.
35 Trinidad and Tobago    Americas   2007    69.8  1056608    18009.
36 West Bank and Gaza     Asia       1972    56.5  1089572     3133.

```{r}
gapminder%>%
  group_by(country)%>%
  arrange(country, year)%>%
  filter(pop<lag(pop) & lifeExp>lag(lifeExp))%>%
  print(n=36)
```
  

