# pre-class


Make sure you commit this often with meaningful messages. 



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


1. Read the source code for each of the following three functions, puzzle out what they do, and then brainstorm better names.

```{r}
#this function takes a string and checks whether or not the prefix (the second argument) exists. It starts at position 1 and checks if the characters in the string match those in the prefix. 
check_if_prefix()<- function(string, prefix) {
  substr(string, 1, nchar(prefix)) == prefix
}

#test f1
f1("yeet", "ye")
f1("yeet", "yo")

#this function deletes the last observation in a given x; it returns null if there is only one observation since then x would be empty
delete_last <- function(x) {
  if (length(x) <= 1) return(NULL)
  x[-length(x)]
}
#testing f2
f2(1)
x<-c(1,2,3,4,5)
f2(x)

#this function takes y and repeats it for each observation in x.
repeat_value<- function(x, y) {
  rep(y, length.out = length(x))
}

#testing f3
f3(x, 9)
f3(9, x)
```

2. Compare and contrast rnorm() and MASS::mvrnorm(). How could you make them more consistent?
```{r}
rnorm
MASS::mvrnorm

#rnorm takes the arguments n, mean, and SD, while mass::mvnorm takes n, mean, and variance rather than SD. We could realign the arguments such that both take SD, for example.
```
3. Use `lapply()` and an anonymous function to find the coefficient of variation (the standard deviation divided by the mean) for all columns in the mtcars dataset. 
```{r}

```
4. Use vapply() to:
    a. Compute the standard deviation of every column in a numeric data frame.
    b. Compute the standard deviation of every numeric column in a mixed data frame. (Hint: you’ll need to use vapply() twice.)
